---
title: "Poisson Modelling within the English Premier League"
author:
  - "Samuel Bleechmore"
  - "Timothy Anderson"
  - "Corey Pickett"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
header-includes:
   - \usepackage[round]{natbib}
   - \usepackage{ifthen}
   - \usepackage{attachfile2}
   - \usepackage{tikz}
   - \usepackage{xstring}
   - \usepackage{enumitem}
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usetikzlibrary{snakes}
   - \usetikzlibrary{trees}
   - \usetikzlibrary{matrix,arrows,decorations.pathmorphing}
   - \usetikzlibrary{decorations.pathreplacing,calc}
   - \usetikzlibrary{shapes}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
options(repos="https://cran.rstudio.com")
```

# Overview

Soccer is a globally popular sport in which two teams compete to score
the most goals within a 90-minute match. Each team fields eleven
players, and goals are relatively rare events compared to other aspects
of gameplay such as passes, tackles, and shots. Because goals occur
infrequently and largely independently of one another, it is common to
model them using a Poisson process, where the number of goals scored in
a fixed time interval follows a Poisson distribution. This modelling
framework assumes that events occur independently and at a constant
average rate, making it suitable for describing goal-scoring patterns in
soccer matches.

The English Premier League (EPL) is one of the most competitive and
widely analysed soccer leagues in the world. Statistical models such as
the Poisson distribution are frequently used in sports analytics to
evaluate team performance, predict match outcomes, and inform betting
markets. Understanding whether the number of goals scored per match
follows a Poisson distribution can provide valuable insights into the
underlying dynamics of scoring and help determine whether assumptions
made by predictive models are valid.

In this project, we will explore whether the distribution of goals
scored by teams in the 2023--24 EPL season is well-approximated by a
Poisson model. Specifically, we will compare the observed distribution
of goals per game with theoretical Poisson distributions and investigate
whether home and away teams exhibit different scoring rates. Given the
well-documented phenomenon of home-field advantage in soccer, we expect
the mean number of goals per game (the Poisson rate parameter,
$\lambda$) to be higher for home teams than for away teams.

\newpage

# The Data

The **worldfootballR** package provides programmatic access to
comprehensive football (soccer) data from FBref.com , including match
results, player statistics, and competition-level summaries across
multiple leagues and seasons. For this project, we extracted data from
the 2023--24 English Premier League (EPL) men's competition using the
package's built-in scraping functions.

The data collection process began by scraping all match URLs for the
target season using the `fb_match_urls()` function. These URLs were then
used to retrieve detailed match results through the `fb_match_urls()`
function. To ensure data quality, matches with missing goal information
were excluded. From the cleaned dataset, a random sample of 50 matches
was drawn for analysis to maintain a manageable yet representative
subset of games.

The data can be obtained and imported into `R` with the following code:

```{r}
library(worldfootballR)
library(dplyr)
library(tidyverse)
library(ggplot2)

# Scraping the match URLs for the EPL 2023-24, men's competition
match_urls = fb_match_urls(
  country = "ENG",
  gender = "M",
  season_end_year = 2024,
  tier = "1st"
)

# Using the scraped match URLs to get match data
match_data = fb_match_results(
  country = "ENG",
  gender = "M",
  season_end_year = 2024,
  tier = "1st"
)

# Removing any matches with missing data
match_data_clean = filter(match_data, !is.na(HomeGoals) & !is.na(AwayGoals))

# Randomly sampling 50 matches
sample_50_matches = sample_n(match_data_clean, 50)

# Converting dataset into long format for easier data analysis
# Will give 1 row per team per match
sample_long_select = select(sample_50_matches, Date, Home, Away, HomeGoals, AwayGoals)
sample_long_select

sample_long_pivot = pivot_longer(
  sample_long_select,
  cols = c(HomeGoals, AwayGoals),
  names_to = "side",
  values_to = "goals"
)

sample_long_mutate = mutate(
  sample_long_pivot,
  team_side = if_else(side == "HomeGoals", "Home", "Away")
)

# Display relevant features
sample_long = select(
  sample_long_mutate,
  Date,
  match_id = Home,
  opponent = Away,
  team_side,
  goals
)

# Home Goals for sample
home_goals = match_data_clean$HomeGoals
# Away Goals for sample
away_goals = match_data_clean$AwayGoals
```

# Topics Covered

The STAT2300 topics covered in this project include:

\begin{itemize}

\item \textbf{Maximum likelihood estimation (MLE)} --- the number of
goals scored by a team in a match can be reasonably modelled by a
Poisson distribution; we therefore use maximum likelihood estimation to
estimate the rate parameter $\lambda$, representing the average goals
scored per match for home and away teams in the 2023--24 English Premier
League season.

\item \textbf{QQ plot and Kolmogorov–Smirnov test} --- we use QQ plots
to visually assess whether it is reasonable to assume that goal counts
follow a Poisson distribution, and apply the Kolmogorov--Smirnov test to
formally evaluate the goodness of fit between the empirical and
theoretical Poisson distributions.

\item \textbf{Confidence interval using the Newton–Raphson method} ---
after verifying that the Poisson distribution adequately describes the
data, we employ the Newton--Raphson procedure to numerically determine
the limits of a likelihood-based confidence interval for the rate
parameter $\lambda$, providing a 95% likelihood interval for the
expected number of goals scored per match.

\end{itemize}

# Results

## Maximum likelihood estimation

The Poisson distribution is a discrete probability distribution commonly
used to model the number of events occurring in a fixed interval of time
or space when these events occur independently and at a constant rate.
In our context, the number of goals scored by a team in a football match
can reasonably be assumed to follow a Poisson distribution, provided
scoring events are independent and occur at a roughly constant rate
throughout the match.

For a Poisson random variable $X$ with rate parameter $\lambda >0$, the
probability mass function (PMF) is given by:

$$
f(x; \lambda) = \frac{e^{-\lambda} \lambda^x}{x!}, \quad x = 0, 1, 2, \ldots
$$

We can use maximum likelihood estimation (MLE) to estimate $\lambda$,
representing the average number of goals scored per match for home and
away teams, respectively.

The likelihood function for a random sample $X= (X_1, X_2,...,X_n)$ is:

$$
\begin{aligned}\mathcal{L}(\lambda|\mathbf{X}) &= \prod_{i = 1}^n \frac{e^{-\lambda}\lambda^{X_i}}{X_i!}\\&= e^{-n\lambda}\lambda^{\sum_{i=1}^n X_i}\left(\prod_{i=1}^n X_i!\right)^{-1}\end{aligned}
$$

Taking the natural logarithm gives the log-likelihood function:

$$
\begin{aligned}\ell(\lambda|\mathbf{X}) &= \ln \mathcal{L}(\lambda|\mathbf{X})\\&= -n\lambda + \left(\sum_{i=1}^n X_i\right)\ln(\lambda) - \sum_{i=1}^n \ln(X_i!)\end{aligned}
$$

To find the MLE of $\lambda$, we differentiate the log-likelihood with
respect to $\lambda$ and set it equal to zero:

$$
\begin{aligned}\frac{\partial}{\partial \lambda}\ell(\lambda|\mathbf{X}) &= -n + \frac{\sum_{i=1}^n X_i}{\lambda} = 0\\\Rightarrow \hat{\lambda} &= \frac{1}{n}\sum_{i=1}^n X_i = \overline{X}\end{aligned}
$$

To confirm that this solution corresponds to a maximum, we compute the
second derivative:

$$
\frac{\partial^2}{\partial \lambda^2}\ell(\lambda|\mathbf{X}) = -\frac{\sum_{i=1}^n X_i}{\lambda^2}
$$

Since this quantity is negative for all $\lambda > 0$, the
log-likelihood is concave and $\hat{\lambda}$ is indeed the MLE.

Hence, the maximum likelihood estimator for the Poisson rate parameter
is:

$$
\boxed{\hat{\lambda} = \overline{X}}
$$

That is, the estimated rate parameter equals the sample mean of observed
goals.

For our data, the estimates can be computed separately for home and away
goals as follows:

```{r}
# Home team goals
lambda_home = mean(sample_long$goals[sample_long$team_side == "Home"])
lambda_home

# Away team goals
lambda_away = mean(sample_long$goals[sample_long$team_side == "Away"])
lambda_away
```

These estimates represent the
\textbf{average number of goals scored per match} by home and away
teams, respectively, during the 2023--24 English Premier League season.
As our results show
$\hat{\lambda}_{\text{home}} > \hat{\lambda}_{\text{away}}$, this
supports the well-established \textit{home advantage phenomenon} in
football.

## QQ plot and Kolmogorov-Smirnov test

A QQ plot can be used to provide a visual assessment of whether it may
be reasonable to assume that the number of goals scored by teams in a
match follows a Poisson distribution. In this context, the Poisson model
assumes that goals occur independently and at a constant rate over the
duration of a match.

There are several ways to produce a QQ plot in `R`. Here, we compare the
empirical quantiles of the observed data to the theoretical quantiles of
a Poisson distribution with parameter values estimated via maximum
likelihood:

$\hat{\lambda}_{\text{home}} = `r round(lambda_home, 3)`$ for home goals
and

$\hat{\lambda}_{\text{away}} = `r round(lambda_away, 3)`$ for away
goals.

```{r qq_home, fig.cap = "QQ Plot: Home Goals vs Poisson"}
# QQ-Plot for Home Goals

ggplot(data.frame(

  x = qpois(ppoints(length(home_goals)), lambda_home), 

  y = sort(home_goals)), 

  aes(x, y)) +

  geom_point() +

  geom_abline(slope = 1, intercept = 0, color = "red") +

  labs(title = "QQ Plot: Home Goals vs Poisson",

       x = "Theoretical Quantiles",

       y = "Observed Home Goals")
```

```{r qq_away, fig.cap = "QQ Plot: Away Goals vs Poisson"}
# QQ-Plot for Away Goals

ggplot(data.frame(

  x = qpois(ppoints(length(away_goals)), lambda_away), 

  y = sort(away_goals)), 

  aes(x, y)) +

  geom_point() +

  geom_abline(slope = 1, intercept = 0, color = "red") +

  labs(title = "QQ Plot: Away Goals vs Poisson",

       x = "Theoretical Quantiles",

       y = "Observed Away Goals")


```

The points in both QQ plots generally align with the $y = x$ reference
line, suggesting that the Poisson model provides a reasonable
approximation for the goal counts. The alignment is slightly tighter for
home goals, consistent with the lower variability observed among home
team scoring outcomes.

To formally test the adequacy of the Poisson model, we use the
Kolmogorov--Smirnov (KS) test implemented through the `ks.test()`
function in `R`. This test compares the empirical cumulative
distribution function (CDF) of the data with the theoretical CDF under a
Poisson distribution with mean parameter $\hat{\lambda}$.

```{r}
# KS Test for Home Goals

ks.test(home_goals, "ppois", lambda_home)

# KS Test for Away Goals

ks.test(away_goals, "ppois", lambda_away)

```

Both tests return very small $p$-values ($p < 2.2 \times 10^{-16}$),
indicating statistically significant deviations between the empirical
and theoretical distributions. However, it is important to note that the
KS test assumes a continuous reference distribution, and its application
to discrete data such as Poisson counts can yield misleadingly small
$p$-values and warnings regarding ties in the data.

As an alternative, we apply a chi-squared goodness-of-fit test, which is
more appropriate for discrete count data and can avoid issues with tied
values. The expected frequencies are computed under the Poisson model
using the MLE $\hat{\lambda}_{\text{home}}$:

```{r}
observed <- table(home_goals)
expected <- dpois(as.numeric(names(observed)), lambda_home) * length(home_goals)
chisq.test(x = observed, p = expected, rescale.p = TRUE)
```

The chi-squared test yields a test statistic of $\chi^2 = 4.622$ with
$6$ degrees of freedom and a $p$-value of $p = 0.593$, indicating no
evidence to reject the null hypothesis that home goals follow a Poisson
distribution. Taken together with the QQ plot, this suggests that the
Poisson model is a reasonable fit for the observed scoring data.

Overall, the QQ plots and chi-squared test results support the
assumption that both home and away goals in the 2023--24 EPL season can
be modelled as Poisson random variables with rate parameters
$\hat{\lambda}_{\text{home}}$ and $\hat{\lambda}_{\text{away}}$
estimated by maximum likelihood.

## Confidence Interval

Our estimates of the average goals per match for home and away teams,
$\hat{\lambda}_{\text{home}} = 1.92$ and
$\hat{\lambda}_{\text{away}} = 1.42$, are point estimates obtained via
maximum likelihood. To provide a range of plausible values for
$\lambda$, we construct 95% likelihood-based confidence intervals using
the Newton--Raphson method.

As derived earlier, the relative likelihood function for the Poisson
rate parameter is: $$
R(\lambda) = \frac{\mathcal{L}(\lambda|\mathbf{X})}{\mathcal{L}(\hat{\lambda}|\mathbf{X})}
= e^{-n(\lambda - \bar{X})}\left(\frac{\lambda}{\bar{X}}\right)^{n\bar{X}}.
$$ The corresponding relative log-likelihood function is therefore: $$
r(\lambda) = \ln R(\lambda) = -n(\lambda - \bar{X}) + n\bar{X}[\ln(\lambda) - \ln(\bar{X})].
$$

Differentiating with respect to $\lambda$ gives: $$
\frac{dr(\lambda)}{d\lambda} = -n + \frac{n\bar{X}}{\lambda}.
$$

The Newton--Raphson algorithm can then be used to iteratively solve for
values of $\lambda$ that satisfy $r(\lambda) = \ln(p)$, where $p$ is the
likelihood ratio cutoff corresponding to a $100C\%$ confidence interval.
For a 95% likelihood interval, $p$ can be calculated as: $$
p = \exp\left\{-\frac{1}{2}\Phi^{-1}\left(\frac{1 - C}{2}\right)^2\right\}
= \exp\left\{-\frac{1}{2}\Phi^{-1}(0.025)^2\right\}
\approx 0.1465.
$$

In `R`, the algorithm was implemented as follows:

```{r}
LI_Poisson = function(initial_lambda, x, thresh = 4, p = 0.95, breakthresh = 5,
                      maxiterations = 1000, print_to_screen = TRUE, makeplot = TRUE)
{
  xbar = mean(x) # sample mean
  n = length(x) # sample size
  counter = 0 # iteration counter
  old_lambda = initial_lambda
  lambda_vec = c(old_lambda)
  diff_lambda = 1

  while (diff_lambda > 10^(-thresh)) {
    # Relative log-likelihood and its derivative
    r_lambda = -n*(old_lambda - xbar) + n*xbar*(log(old_lambda) - log(xbar))
    deriv = -n + n*xbar/old_lambda
    new_lambda = old_lambda - (r_lambda - log(p))/deriv

    lambda_vec = c(lambda_vec, new_lambda)
    counter = counter + 1
    if (print_to_screen == TRUE) print(c(counter, new_lambda))
    diff_lambda = abs(old_lambda - new_lambda)

    if (diff_lambda > 10^breakthresh | counter >= maxiterations) break
    old_lambda = new_lambda
  }

  if (makeplot == TRUE) plot(lambda_vec, type = "o", xlab = "Iteration", ylab = expression(hat(lambda)))
  converged = diff_lambda < 10^(-thresh)
  return(list(LI_NR = new_lambda, last_iteration_diff = diff_lambda, converged = converged))
}
```

To construct the confidence interval, we provide initial guesses below
and above the MLE $\hat{\lambda}$ to estimate the lower and upper limits
of the interval. The function used to convert a likelihood estimate into
a confidence interval is below.

```{r}
# Conversion Function
LikePFromConf = function(C = 0.95)
{
  exp(-1/2*qnorm((1 - C)/2)^2)
}
LikePFromConf(C = 0.95)
```

For the home goals data:

```{r}


LL95_Home = LI_Poisson(initial_lambda = 0.5*mean(home_goals), x = home_goals, 
                       p = LikePFromConf(0.95))

UL95_Home = LI_Poisson(initial_lambda = 2*mean(home_goals), x = home_goals, 
                       p = LikePFromConf(0.95))

```

The algorithm converged after 7 iterations for the upper limit and 6 for
the lower limit, yielding: The 95% likelihood interval for home goals is
$(1.668, 1.938)$.

Similarly, for the away goals data:

```{r}

LL95_Away = LI_Poisson(initial_lambda = 0.5*mean(away_goals), x = away_goals, 
                       p = LikePFromConf(0.95))

UL95_Away = LI_Poisson(initial_lambda = 2*mean(away_goals), x = away_goals, 
                       p = LikePFromConf(0.95))

```

The algorithm converged after 6 iterations for both limits, giving: The
95% likelihood interval for away goals is $(1.360, 1.605)$.

Both sets of intervals indicate that the average number of goals scored
per match by home teams is slightly higher than that of away teams. The
relatively narrow intervals suggest that the estimates are precise, and
the overlap between home and away goal intervals implies that while a
home advantage exists, its magnitude is moderate.

# Conclusion

This project set out to determine whether the distribution of goals
scored by teams in the 2023--24 season of the English Premier League is
well approximated by a Poisson distribution. This was done using various
techniques from STAT2300. The Poisson model was chosen because goal
scoring in soccer occurs largely independently and at an approximately
constant rate over the duration of a match.

Using maximum likelihood estimation, we obtained rate parameter
estimates of:

$$
\hat{\lambda}_{\text{home}} = 1.92 \quad \text{and} \quad \hat{\lambda}_{\text{away}} = 1.42
$$

This result is consistent with the well established home-field
advantage, where home teams typically score more goals on average than
away teams.

Visual assessment through QQ plots showed good alignment with the
$y = x$ reference line, especially for home goals, indicating that the
Poisson model provides a good approximation for the observed goal
counts.

The Kolmogorov--Smirnov test produced very small $p$-values
($p < 2.2 \times 10^{-16}$), indicating apparent deviations between the
empirical and theoretical distributions. However, this result may be
misleading, as the Kolmogorov--Smirnov test assumes a continuous
reference distribution, whereas our data is discrete.

Therefore, the chi-squared goodness-of-fit test was performed instead,
producing a $p$-value of 0.593. This provided no evidence to reject the
null hypothesis that the data follows a Poisson distribution.

We constructed 95% confidence intervals using the Newton–Raphson method.
The interval for home goals was (1.668, 1.938), while the interval for
away goals was (1.360, 1.605). These relatively narrow intervals suggest
that our estimates are precise and stable. Additionally, the fact that
the home interval lies consistently above the away interval provides
support for the presence of home advantage, while the moderate overlap
reflects the competitive balance across teams. Together, these intervals
strengthen the conclusion that the Poisson model not only fits the data
well but also yields reliable estimates of scoring behavior in the EPL.
